Sulfur Dioxide on Water's Surface

### 2011-06-24 Conclusions.tex
Draft 7

% intro-type paragraph here
The aqueous chemistry taking place in the Earth's atmosphere is goverened by the physics of the water surfaces found there. Many recent experiments have revealed specific pathways for reaction of \suldiox, but the molecular-level nature of those reactions are still poorly understood. Through computer simulations we are able to probe the microscopic environments in which those reactions occur, and better understand the conditions in which \suldiox~reacts.

@@

Our simulations recreated the experimental setup of a recent study by our group,\cite{Ota2011} by introducing a concentrated gas phase of \suldiox~to a water surface. Through simulation it was found that gaseous \suldiox~quickly adsorbs to the water surface, and continues to bind until a complete surface coverage is reached. Surface waters reorient in the presence of adsorbed \suldiox. The waters at the interface of a neat-water surface tend to lay flatter to the surface than when a saturating layer of \suldiox~is present. The waters interacting with the layer of adsorbed \suldiox~orient more perpendicularly to the interface, and further expose their ``free-OH'' uncoupled bonds for interactions with \suldiox, and hydrate complex formation.

@@

Steered molecular dynamics simulations were used to model the behavior of an adsorbing \suldiox~as it moves from the gas phase above the water down through the surface and into the bulk. The results for the transit through the interface show that in both systems of low and high \suldiox~concentration an adsorbing \suldiox~has very similar orientation to those already bound to the water surface. As the \suldiox~approaches the water surface the first contact with water causes it to reorient. Within 5 $\AA$ of the surface the \suldiox~is mostly oriented with its sulfur towards the water phase. The \suldiox~pulled further into the water bulk retains its orientation until it is past the interfacial region and then isotropically orients with the bulk water.

This study is one of several aiming to characterize \suldiox~adsorption and behavior on aqueous surfaces. The atmospheric chemistry driving many important reactions occurs on aerosol and small particle surfaces. To gain a better understanding of gaseous adsorption we plan to report further computational simulation and experimental results of temperature and chemical constituent effects on atmospherically relevant water surfaces.

>>>>

% intro-type paragraph here
The aqueous chemistry taking place in the Earth's atmosphere is governed by the physics and chemistry of the water surfaces found there. Many recent experiments have revealed specific pathways for reaction of \suldiox, but the molecular-level nature of those reactions are still poorly understood. Through computer simulations we are able to probe the microscopic environments in which those reactions occur, and better understand the conditions in which \suldiox~reacts.

@@

Our simulations recreated the experimental setup of a recent study by our group,\cite{Ota2011} by introducing a concentrated gas phase of \suldiox~to a water surface. Through simulation it was found that gaseous \suldiox~quickly adsorbs to the water surface, and continues to bind until a complete surface coverage is reached. Surface waters reorient in the presence of adsorbed \suldiox. The waters at the interface of a neat-water surface tend to lay flatter to the surface than when a saturating layer of \suldiox~is present. The waters interacting with the layer of adsorbed \suldiox~orient more perpendicularly to the interface, and further expose their ``free-OH'' uncoupled bonds for interactions with \suldiox, and hydrate complex formation. Furthermore, we have found that surface waters underneath a blanket layer of adsorbed \suldiox~will penetrate further into the gas phase, allowing for greater mobility of waters away from the aqueous bulk.

@@

Steered molecular dynamics simulations were used to model the behavior of an adsorbing \suldiox~as it moves from the gas phase above the water down through the surface and into the bulk. The results for the transit through the interface show that in both systems of low and high \suldiox~concentration an adsorbing \suldiox~has very similar orientation to those already bound to the water surface. The \suldiox~reorients as it makes its first contact with the water surface. Within 5 $\AA$ of the surface the \suldiox~is mostly oriented with its sulfur towards the water phase. The \suldiox~pulled further into the water bulk retains its orientation until it is past the interfacial region and then isotropically orients with the bulk water.

Having now examined this passageway of \suldiox~from the gas to adsorbed aqueous phases, we may now turn to further filling in the details of adsorption and interfacial chemistry of gas molecules. What species form during the adsorption transition, and how do molecular properties affect the process? Can we form a theory that will more fully explain the transit of many small molecules of environmental and industrial interest? This study is one of several aiming to characterize \suldiox~adsorption and behavior on aqueous surfaces. We plan to report further results of computational simulation and experimental studies regarding temperature and chemical constituent effects on atmospherically relevant water surfaces.

### 2011-06-24 DensityResults.tex
Draft 7

One measure of surface activity is the spatial distribution of molecules in the interfacial region. The density distributions of both \wat~and \suldiox~were calculated for the equilibrium MD simulations. The results presented in figure \ref{fig:density} show both the water and \suldiox~density distributions across the two simulated slabs. As shown, the \suldiox~exposed to the neat-water system remained at the water slab surface. The \suldiox~in the saturated system accumulated mostly at the surface, but some residual \suldiox~remained well into the bulk water.

@@

In the simulated neat-water slab, the \suldiox~molecular density distribution concentrates near the water surface, indicating an affinity for the interfacial region. This is consistent with what is found experimentally in our laboratory using VSFS,\cite{Tarbuck2005,Tarbuck2006} and also verified by the computational simulations and spectral calculations of Baer et al.\cite{Baer2010} The experimental studies indicated that upon exposure of \suldiox~to a \wat~surface, a layer of \suldiox~hydrate complex forms, modifying the structure of water above the Gibbs dividing surface (GDS). In the experimental system, the \suldiox~bonding interaction with the free OH oscillators is manifested in a red-shift of the free OH frequency indicative of a bonding interaction. In a later section of this article we look at the effect of \suldiox~on surface water orientation.  During the course of the simulation, the \suldiox~does not venture into the bulk water. Also, the \suldiox~does not escape the water surface into the gas phase, but remains located within 5-\angs~of the surface.

>>>>

One measure of surface activity is the spatial distribution of molecules in the interfacial region. The density distributions of both \wat~and \suldiox~were calculated for the equilibrium MD simulations. The results presented in figure \ref{fig:density} show both the water (black) and \suldiox~(red) density distributions across the two simulated slabs. As shown, the \suldiox~exposed to the neat-water system remained at the water slab surface. The \suldiox~in the saturated system accumulated mostly at the surface, but some residual \suldiox~remained well into the bulk water.

@@

In the simulated neat-water slab, the result that the \suldiox~molecular density distribution concentrates near the water surface indicates an affinity for the interfacial region. This is consistent with what is found experimentally in our laboratory using VSFS,\cite{Tarbuck2005,Tarbuck2006} and also supported by the computational simulations and spectral calculations of Baer et al.\cite{Baer2010} The experimental studies indicated that upon exposure of \suldiox~to a \wat~surface, a layer of \suldiox~hydrate complex forms, modifying the structure of water above the Gibbs dividing surface (GDS). In the experimental system, the \suldiox~bonding interaction with the free OH oscillators is manifested in a red-shift of the free OH frequency indicative of a bonding interaction. In a later section of this article we look at the effect of \suldiox~on surface water orientation.  During the course of the simulation, the \suldiox~does not venture into the bulk water nor does it escape the water surface into the gas phase, but remains located within 5-\angs~of the surface region.

### 2011-06-24 EquilibratedOrientation.tex
Draft 7

The bisector tilt of the water molecules, $\cos(\theta)$, concentrates around $\cos(\theta)=0$ within the first few\angs~above and below the surface, becoming progressively isotropic further into the water bulk of both systems. As the tilt nears $\cos(\theta)=0$ the \wat~bisector lies within the plane of the surface indicating a water orientation either flat on the surface, or with some amount of ``twist'' sending the OH bonds in towards, or out of the bulk. The value of $\phi$ determines the ``twist'' in this case. Both systems show a defined intensity concentration (dark red) in the distributions around $\cos(\phi)=1$ at the water surfaces. This results from an orientation of the water's y-axis (normal to the molecular plane) aligned perpendicular to the plane of the water surface. Thus, for both the neat-water and saturated systems, the preferred orientation of water molecules at a distance of 0\angs~ is to lie mostly flat to the plane of the interface.

Although the plots show overall similarities for both the neat-water and saturated systems, the presence of a layer of adsorbed \suldiox~molecules alters the water orientation of the waters furthest into the gas phase. The waters located above 0\angs~are further into the gas phase, and interact with the layer of adsorbed \suldiox~molecules. The resulting orientation of waters above 0\angs, shown in the saturated plot of $\cos(\theta)$ of figure \ref{fig:water-orientation}, is with a bisector pointing further into the adsorbed \suldiox~gas layer, and both hydrogens pointing outward from the aqueous bulk. The effect is more pronounced further from the water phase, and above 5\angs~the $\cos(\theta)$ distribution is completely centered around $\cos(\theta)=+1$ (see figure \ref{fig:water-angles}).

@@

The noise in the neat-water plots of figure \ref{fig:water-orientation} above 5\angs~(manifested as disconnected points of intensity throughout the range of orientations) is a result of fewer waters venturing beyond those extents and thus less data far from the surface. This is one indication that waters near a layer of adsorbed \suldiox~venture further above the interface and interact with the adsorbed gas molecules. The interactions with neighboring \suldiox~molecules allow the waters above the surface to orient more perpendicularly to the interface. Our experimental SFG studies also predicted this reorienting behavior due to the \suldiox~interactions with the topmost surface waters.\cite{Ota2011}

>>>>

The bisector tilt of the water molecules, $\cos(\theta)$, concentrates around $\cos(\theta)=0$ within the first few\angs~above and below the surface, becoming progressively isotropic further into the water bulk of both systems. As the tilt nears $\cos(\theta)=0$ the \wat~bisector lies within the plane of the surface indicating a water orientation either flat on the surface, or with some amount of ``twist'' sending the OH bonds in towards, or out of the bulk. The value of $\phi$ determines the ``twist'' in this case. Both systems show a defined intensity concentration (dark red) in the distributions around $\cos(\phi)=1$ at the water surfaces. This results from an orientation of the water's y-axis (normal to the molecular plane) aligned perpendicular to the plane of the water surface. Thus, for both the neat-water and saturated systems, the preferred orientation of water molecules at a distance of 0\angs~ is to lie mostly flat to the plane of the interface with a slight ``twist'' sending one OH bond further into the gas phase, and the other OH towards the water bulk. This result agrees somewhat with a recent air/water study by Fan et al. in which the surface orientations of several water models were analyzed.\cite{Fan2009} They simulated water using non-polarizable models, however, which alters the behavior of water at interfacial regions when compared to the polarizable POL3 model used in our work. The main conclusions are similar in that one OH tends to point further into the air phase than the other, but differs in that this effect is less pronounced with the polarizable model.

Although the plots show overall similarities for both the neat-water and saturated systems, the presence of a layer of adsorbed \suldiox~molecules alters the water orientation of the waters furthest into the gas phase. For the saturated solution the resulting orientation of waters above 0\angs, shown in the saturated plot of $\cos(\theta)$ of figure \ref{fig:water-orientation}, is with a bisector pointing further into the adsorbed \suldiox~gas layer, and both hydrogens pointing outward from the aqueous bulk. The effect is more pronounced further from the water phase, and above 5\angs~the $\cos(\theta)$ distribution is completely centered around $\cos(\theta)=+1$ (see figure \ref{fig:water-angles}).

@@

The noise in the neat-water plots of figure \ref{fig:water-orientation} above 5\angs~(manifested as disconnected points of intensity throughout the range of orientations) is a result of fewer waters venturing beyond those extents and thus less data far from the surface. This is one indication that waters near a layer of adsorbed \suldiox~venture further above the interface relative to the low \suldiox~concentration, where they can have more \suldiox~interaction with the adsorbed gas molecules. These results show that the interactions with neighboring \suldiox~molecules allow the waters above the surface to orient more perpendicularly to the interface. This is consistent with our recent experimental VSFS studies which showed evidence for the reorienting behavior of water due to the \suldiox~interactions with the topmost surface waters.\cite{Ota2011}

### 2011-06-24 Introduction.tex
Draft 7

The doorway to the uptake of a gas by an aqueous solution is the water surface. Although we know much about the behavior of a gas on either side of that entrance, far less is known about how that surface acts to attract, facilitate, or thwart the transit of the gas between the two bulk phases. What species form during gas adsorption onto liquid surfaces, and what are the intermediary steps?  To what extent does an adsorbing gas affect liquid water surface molecules, or the structure of the interface throughout an adsorption process?  Are specific gas or liquid molecular orientations necessary for gaseous adsorption?  Experimental studies to address such questions are valuable but do not provide the resolution necessary to determine the geometries of adsorbing gases, or to determine the orientations of the molecules at the liquid surface near the adsorption site.
This type of information can be determined computationally, and when coupled to the experimental studies can provide important new insight into the gas-liquid surface adsorption process.

In this work we look at the orientational properties of sulfur dioxide gas adsorbing to a water surface as a model for the transit of gases across an aqueous surface.  The interactions between sulfur species and aqueous systems are currently an active area of chemical research. Sulfur dioxide enters the environment as an important industrial product, and also naturally through terrestrial processes. Atmospheric dust particles and gases have been implicated in the oxidation of \suldiox, and act as reaction surfaces for chemical mechanisms that are still poorly understood.\cite{Baltrusaitis2011,Rubasinghege2010,Li2007} \suldiox~acts as a major component of atmospheric pollution, and is a precursor to acid rain formation, and cloud nucleation. Its high solubility in water makes \suldiox~an integral compound in many aqueous atmospheric reactions, as well. We seek a fuller understanding of the interactions between \suldiox~and water at interfaces because they are central to many atmospheric particle and aerosol reactions.

@@

Our latest VSFS experiments on \suldiox~behavior further expand our understanding of \suldiox~in atmospheric processes.\cite{Ota2011} A temperature study has shown that the binding of gaseous \suldiox~to water surfaces is greatly enhanced at atmospherically relevant cold temperatures. The surface binding of unreacted \suldiox~complexes is found to be a completely reversible process. The same study also showed that low pH aqueous environments inhibit the bulk reaction of \suldiox, but do not affect the surface binding of the gas molecules. Furthermore, the resultant experimental spectra suggest that the surface waters reorient in response to the presence of adsorbed \suldiox, but the details about this reorientation can not be garnered from the experiments.  Such details include knowing how surface water orientation changes upon approach of the gaseous \suldiox, how this varies upon \suldiox~surface bonding, and the orientation of \suldiox~throughout the surface bonding process. 
%Knowing that reorientation may occur due to \suldiox~adsorption, how can we confirm it? what is the nature of the reorienting waters? How do they reorient, and to what extent? Conclusive experimental evidence of the adsorbing gas or surface water molecular orientation is still missing, so we are moved to use computational methods for further understanding this phenomena.

We provide such detailed information to complement and augment these experimental studies. Our studies involve computation of the orientational depth profiles of water and \suldiox~before and after the adsorption process. They augment the spectroscopic experiments that can not probe depth or orientation profiles of surface species in the same detailed manner as computational simulations. The molecular behaviors that occur upon introduction of \suldiox~to a \wat~surface are determined, and we analyze the orientational response of both the \suldiox~and \wat~molecules using equilibrium and steered (SMD) classical molecular dynamics simulations. The latter approach involves steering a gas molecule into the aqueous phase, and characterizing its molecular orientation as it transits through the interfacial region.  This unique approach enables new insights into the behaviors of gas molecules as they move near to liquid water to be obtained. We also simulate how \suldiox~adsorption occurs on a water surface saturated with adsorbed \suldiox, analogous to the cold \suldiox~experimental studies recently performed.\cite{Ota2011} The results of this study provide an intimate perspective on the adsorption of \suldiox~at an aqueous surface, and a more complete picture of gaseous adsorption to liquid interfaces.

>>>>

The doorway to the uptake of a gas by an aqueous solution is the water surface. Although we know much about the behavior of a gas on either side of that entrance, far less is known about how that surface acts to attract, facilitate, or thwart the transit of the gas between the two bulk phases. What species form during gas adsorption onto liquid surfaces, and what are the intermediary steps?  To what extent does an adsorbing gas affect liquid water surface molecules, or the structure of the interface throughout an adsorption process?  Are specific gas or liquid molecular orientations necessary for gaseous adsorption?  Experimental studies to address such questions are valuable but do not provide the full resolution necessary to determine the geometries of adsorbing gases, or to determine the orientations of the molecules at the liquid surface near the adsorption site.
This type of information can be determined computationally, and when coupled to the experimental studies can provide a more comprehensive picture of the gas-liquid surface adsorption process.

In this work we look at the orientational properties of sulfur dioxide gas adsorbing to a water surface as a model for the transit of gases across an aqueous surface.  The interactions between sulfur species and aqueous systems are currently an active area of chemical research and general interest.\cite{Lattanzi2010,Shah2011,Tzivian2011,Johns2011,Faloona2009,Jurkat2010,Wu2011} Sulfur dioxide enters the environment as an important industrial product, and also naturally through terrestrial processes. Atmospheric dust particles and gases have been implicated in the oxidation of \suldiox, and act as reaction surfaces for chemical mechanisms that are still poorly understood.\cite{Baltrusaitis2011,Rubasinghege2010,Li2007} \suldiox~acts as a major component of atmospheric pollution, and is a precursor to acid rain formation, and cloud nucleation. Its high solubility in water makes \suldiox~an integral compound in many aqueous atmospheric reactions, as well. We seek a fuller understanding of the interactions between \suldiox~and water at interfaces because they are central to many atmospheric particle and aerosol reactions.

@@

Our latest VSFS experiments on \suldiox~behavior further expand our understanding of \suldiox~in atmospheric processes.\cite{Ota2011} A temperature dependent study has shown that the binding of gaseous \suldiox~to water surfaces is greatly enhanced at atmospherically relevant cold temperatures. The surface binding of unreacted \suldiox~complexes is found to be a completely reversible process when at cold temperatures. The same study also showed that low pH aqueous environments inhibit the bulk reaction of \suldiox, but do not affect the surface binding of the gas molecules. Furthermore, the resultant experimental spectra suggest that the surface waters reorient in response to the presence of adsorbed \suldiox, but the details about this reorientation can not be garnered from the experiments.  Such details include knowing how surface water orientation changes upon approach of the gaseous \suldiox, how this varies upon \suldiox~surface bonding, and the orientation of \suldiox~throughout the surface bonding process. 
%Knowing that reorientation may occur due to \suldiox~adsorption, how can we confirm it? what is the nature of the reorienting waters? How do they reorient, and to what extent? Conclusive experimental evidence of the adsorbing gas or surface water molecular orientation is still missing, so we are moved to use computational methods for further understanding this phenomena.

Here we provide such detailed information to complement and augment these experimental studies. Our studies involve computation of the orientational depth profiles of water and \suldiox~before and after the adsorption process. The molecular behaviors that occur upon introduction of \suldiox~to a \wat~surface are determined, and the orientational response of both the \suldiox~and \wat~molecules using equilibrium and steered (SMD) classical molecular dynamics simulations are analyzed. The latter approach involves steering a gas molecule into the aqueous phase, and characterizing its molecular orientation as it transits through the interfacial region.  This unique approach enables new insights into the behaviors of gas molecules as they move near to liquid water to be obtained. We also simulate how \suldiox~adsorption occurs on a water surface saturated with adsorbed \suldiox, analogous to the cold \suldiox~experimental studies recently performed.\cite{Ota2011} The results of this study provide an intimate perspective on the adsorption of \suldiox~at an aqueous surface, and a more complete picture of gaseous adsorption to liquid interfaces.

### 2011-06-24 Methods.tex
Draft 7

The low and high concentration simulated systems, ``neat-water'' and saturated, respectively, were created as follows: the ``neat-water'' simulation involved the addition of a single \suldiox~molecule either within the bulk of the water slab (for equilibrium MD), or above the slab surface (in the SMD simulations). The single-\suldiox~neat-water system was then evolved for 2 ns to produce an equilibrated starting configuration. Because of the extremely low concentration of \suldiox~in the neat-water system the surface waters behave similarly to a true ``neat-water'' air-liquid interface. The neat-water system orientational results shown later in this work reproduce well the results of our previous orientational studies of surface water behavior.\cite{Walker2006b,Hore2008} The saturated system had 22 \suldiox~molecules introduced to the water slab bulk in order to saturate it to a level coinciding with the Henry's law constant for \suldiox~in water. Additionally, 50 \suldiox~molecules were introduced into the gas phase outside of the saturated water slab to simulate an added \suldiox~gas pressure. The saturated system with both bulk and gaseous \suldiox~was then evolved for 2 ns to produce a starting configuration for further saturated simulations.

\subsection{Equilibrium Simulations}

@@

A separate set of SMD simulations were performed with tethering of one of the \suldiox-oxygens to the water slab center of mass. This was done to ensure that the orientation of the \suldiox~during the adsorption transit was not an artifact of the choice of atom used for tethering. The simulations produced the same results (not shown) for the orientational analyses, so the data from the original tethering scheme were used.

@@

	In this portion of the study we create orientational depth-profiles of \suldiox~and \wat~upon exposure and adsorption of the gas. The reference point used for depth calculations is the location of the water surface. A liquid surface, however, is a dynamic boundary that is neither flat nor stationary. To define a water surface location, several previous studies have used the technique of fitting a line shape to the averaged density profile of the water, and extracting interfacial shape and location parameters.\cite{Shamay2010,Wick2006c,Chowdhary2006} Hyperbolic tangent functions have been used often, and values for the ``Gibb's dividing surface'' location, and interfacial width have thus been determined.\cite{Matsumoto1988} However, in long simulations the location and shape of the interface changes, and the motion of surface waters alters the interfacial width at any given time step. Thus, the density profile fitting will capture averaged widths and locations, not instantaneous values. Similarly, the averaged values of location and width will obscure information about any drift or deformations the surface undergoes.

@@

	In this work we define the surface location by calculating a reference location at each time step by averaging the positions of the waters contained in the topmost monolayer. This provides us with a consistent and intuitive reference point in the simulations to which we relate our analyses, but does not increase the computational burden.

@@

	The simulated systems were analyzed to characterize the orientation of \wat~and \suldiox~in various environments above, within, and below the aqueous surface region. The molecular orientation of both the \wat~and \suldiox~is a prerequisite for understanding the chemistry occurring during the \suldiox~adsorption process. The two molecules studied are similarly shaped with a C2$_v$ axis along their bisectors, and a molecular plane defined by three atoms. A body-fixed frame is defined for both \wat~and \suldiox~as shown in figure \ref{fig:molecular-frame}. In each analysis a space-fixed reference axis is used that corresponds to the long axis of the system's periodic cell normal to the plane of the water surface. The orientational analyses presented herein focus on two angles used to define molecular orientation. The molecular orientation angles $\theta$ and $\phi$ are determined from a set reference axis as shown in figure \ref{fig:water-angles}a.

>>>>

The low and high concentration simulated systems, ``neat-water'' and saturated water, respectively, were created as follows: the neat-water simulation involved the addition of a single \suldiox~molecule either within the bulk of the water slab (for equilibrium MD), or above the slab surface (in the SMD simulations). The single-\suldiox~neat-water system was then evolved for 2 ns to produce an equilibrated starting configuration. Because of the extremely low concentration of \suldiox~in the neat-water system the surface waters behave similarly to a true neat-water air-liquid interface. The neat-water system orientational results shown later in this work reproduce well the results of our previous orientational studies of surface water behavior.\cite{Walker2006b,Hore2008} The saturated system had 22 \suldiox~molecules introduced to the water slab bulk in order to saturate it to a level coinciding with the Henry's law constant for \suldiox~in water. Additionally, 50 \suldiox~molecules were introduced into the gas phase outside of the saturated water slab to simulate an added \suldiox~gas pressure. The saturated system with both bulk and gaseous \suldiox~was then evolved for 2 ns to produce a starting configuration for further saturated simulations.

\subsection{Equilibrium Molecular Dynamics Simulations}

@@

A separate set of SMD simulations were performed with tethering of one of the \suldiox-oxygens to the water slab center of mass. This was done to ensure that the orientation of the \suldiox~during the adsorption transit was not an artifact of the choice of atom used for tethering. The simulations produced the same results (not shown) for the orientational analyses, so the data from the original tethering scheme was used.

@@

	The first portion of the studies involved creating orientational depth-profiles of \suldiox~and \wat~upon exposure and adsorption of the gas. The reference point used for depth calculations is the location of the water surface. A liquid surface, however, is a dynamic boundary that is neither flat nor stationary. To define a water surface location, several previous studies have used the technique of fitting a line shape to the averaged density profile of the water, and extracting interfacial shape and location parameters.\cite{Shamay2010,Wick2006c,Chowdhary2006} Hyperbolic tangent functions have been used often, and values for the ``Gibb's dividing surface'' location, and interfacial width have thus been determined.\cite{Matsumoto1988} However, in long simulations the location and shape of the interface changes, and the motion of surface waters alters the interfacial width at any given time step. Thus, the density profile fitting will capture averaged widths and locations, not instantaneous values. Similarly, the averaged values of location and width will obscure information about any drift or deformations the surface undergoes.

@@

	To overcome this problem we define the surface location by calculating a reference location at each time step by averaging the positions of the waters contained in the topmost monolayer. This provides us with a consistent and intuitive reference point in the simulations to which we relate our analyses, but does not increase the computational burden.

@@

The molecular orientation of both the \wat~and \suldiox~is a prerequisite for understanding the chemistry occurring during the \suldiox~adsorption process. With the surface location as defined above, the simulated systems were analyzed to characterize the orientation of \wat~and \suldiox~in various environments above, within, and below the aqueous surface region. The two molecules studied are similarly shaped with a C2$_v$ axis along their bisectors, and a molecular plane defined by three atoms. A body-fixed frame is defined for both \wat~and \suldiox~as shown in figure \ref{fig:molecular-frame}. In each analysis a space-fixed reference axis is used that corresponds to the long axis of the system's periodic cell normal to the plane of the water surface. The orientational analyses presented herein focus on two angles used to define molecular orientation. The molecular orientation angles $\theta$ and $\phi$ are determined from a set reference axis as shown in figure \ref{fig:water-angles}a.

### 2011-06-24 TransitOrientation.tex
Draft 7

	From its starting position 20\angs~above the water surface, until the \suldiox~moves to within 10\angs~of the surface of both systems, the orientation is isotropic in $\theta$ and $\phi$. This can be seen in the plots as mostly uniform coloration at a given distance from the surface independent of $\cos$ of $\phi$. Near the surface the bisector angle $\theta$ becomes more perpendicular to the water surface ($\cos(\theta)\approx1$) with the \suldiox~sulfur pointing into the water phase. At the point when the \suldiox~reaches the surface location (distance $=0$\angs), the bisector is perpendicular in both the neat-water and saturated systems. Figure \ref{fig:so2-surface-angles}a depicts \suldiox~near the water surface with the bisector aligned perpendicular to the interface. In the absence of simulated ionic species that form through \suldiox-\wat~chemistry at the surface, it is clear that the adsorbing \suldiox~in the gas phase takes on a preferred orientation to adsorb on a water surface. The main difference between the two systems is where the point of the transition from isotropic to preferred orientations is found.

  % talk about theta differences
  Looking more closely at the approach, the $\theta$ distribution above 10\angs~from the surface is isotropic. For the neat-water surface, a transition occurs at approximately 5\angs~from the surface where the \suldiox~orients with its bisector perpendicular to the water surface. This coincides with the approximate distance of bonds formed to the outer-most water molecules. Below 5\angs, the orientation persists and the \suldiox~begins to interact with the topmost surface waters, and orients accordingly. In the saturated system, the same trend occurs, however the onset of the perpendicular orientation begins at approximately 8\angs~above the surface. The layer of adsorbed \suldiox~already present in the saturated system most likely interacts with the transiting \suldiox~molecule. Also, topmost water molecules from the surface move up to a few \angs~inside the adsorbed \suldiox~layer and interact with the transiting \suldiox~further from the surface than those in the neat-water system. It is remarkable that the orientational trend appears so strongly in the $\theta$ plots even with so little data as was collected from the single \suldiox~molecule of each simulation. From onset of orientation above the surface until 10\angs~below, the \suldiox~holds a preferred orientation.

@@

 With a mostly perpendicular bisector angle, it is expected that the values of $\phi$ for \suldiox~would be isotropic relative to the reference axis. This is the case in both systems, with only a few exceptions. At the neat-water surface and just below, from 0\angs~to 5\angs~in the water phase, the $\theta$ profile broadens down to $\cos(\theta)=0.5$, and a clear trend appears in the profile of $\phi$ near $\cos(\phi)=0$. This corresponds to \suldiox~inclined up to 60 degrees from the surface normal. The result is that \suldiox~in this region shows a preference for one of the S-O bonds to point more towards the aqueous bulk and the other to point away towards the interface. Figure \ref{fig:so2-surface-angles}b depicts the orientation of the \suldiox~slightly below the water surface. This is in contrast to \suldiox~molecules above the surface, oriented more perpendicularly. The behavior is likely due to the interaction between the waters and one of the S-O bonds leading to a higher solvation than above the water surface. As the \suldiox~is solvated by more highly-coordinated bulk water, the S-O bonds experience less equal interaction environments. Baer et al. noted that their force field model for the \suldiox~does not reproduce well the first hydration shell geometries,\cite{Baer2010} so conclusions regarding the specific interactions and hydrate geometries between the \suldiox~and \wat~cannot be made here. It is notable that the same reorientation does not occur as strongly in the saturated system. The presence of the adsorbed \suldiox~layer apparently decreases the reorienting behavior likely because of the disrupting effect the higher \suldiox~concentration has on the water interactions at the surface. %Below this 5\angs~region under the surface, the \suldiox~resumes an isotropic orientation as it becomes fully solvated by bulk waters.

>>>>

	From its starting position above the water surface, until the \suldiox~moves to within 10\angs~of the surface of both systems, the orientation is isotropic in $\theta$ and $\phi$ (not shown). Isotropic orientation is manifested in the plots as mostly uniform coloration at a given distance from the surface independent of $\cos$ of $\phi$. Near and into the surface the bisector angle $\theta$ becomes more perpendicular to the water interface ($\cos(\theta)\approx1$) with the \suldiox~sulfur pointing into the water phase, consistent with the equilibrium MD simulation results above. At the point when the \suldiox~reaches the surface location (distance $=0$\angs), the bisector is perpendicular in both the neat-water and saturated systems. Figure \ref{fig:so2-surface-angles}a depicts \suldiox~near the water surface with the bisector aligned perpendicular to the interface. In the absence of simulated ionic species that form through \suldiox-\wat~chemistry at the surface, it is clear that the adsorbing \suldiox~in the gas phase takes on a preferred orientation to adsorb on a water surface. The main difference between the neat and saturated water systems is where the point of the transition from isotropic to preferred orientations is found.

  % talk about theta differences
  Comparing in more detail the difference in these two systems upon \suldiox~approach, for the neat-water surface, a transition occurs at approximately 5\angs~from the surface. This coincides with the approximate distance of bonds formed to the outer-most water molecules. Below 5\angs above the surface, the orientation persists and the \suldiox~begins to interact with the topmost surface waters, and orients accordingly. In the saturated system, the same trend occurs, however the onset of the perpendicular orientation begins at approximately 8\angs~above the surface. The layer of adsorbed \suldiox~already present in the saturated system most likely interacts with the transiting \suldiox~molecule. Also, topmost water molecules from the surface move up to a few \angs~inside the adsorbed \suldiox~layer and interact with the transiting \suldiox~further from the surface than those in the neat-water system. It is remarkable that the orientational trend appears so strongly in the $\theta$ plots even with so little data as was collected from the single \suldiox~molecule of each simulation. From onset of orientation above the surface until 10\angs~below, the \suldiox~holds a preferred orientation.

@@

 With a mostly perpendicular bisector angle, it is expected that the values of $\phi$ for \suldiox~would be isotropic relative to the reference axis. This is the case in both systems, with only a few exceptions. In both systems the $\cos(\phi)$ profiles exhibit mostly isotropic distributions above 0\angs, with several regions of lighter coloration interspersed, but without a clearly formed orientational trend. At the neat-water surface and just below, from 0\angs~to -5\angs~in the water phase, the $\theta$ profile broadens down to $\cos(\theta)=0.5$, coinciding with a clear trend in the profile of $\phi$ near $\cos(\phi)=0$ shown in dark red at the bottom of the plot. This corresponds to \suldiox~inclined up to 60 degrees from the surface normal. The result is that \suldiox~in this region shows a preference for one of the S-O bonds to point more towards the aqueous bulk and the other to point away towards the interface. Figure \ref{fig:so2-surface-angles}b depicts the orientation of the \suldiox~slightly below the water surface. This is in contrast to \suldiox~molecules above the surface, oriented more perpendicularly. The behavior is likely due to the interaction between the waters and one of the S-O bonds leading to a higher solvation than above the water surface. As the \suldiox~is solvated by more highly-coordinated bulk water, the S-O bonds experience less equal interaction environments. Baer et al. noted that their force field model for the \suldiox~does not reproduce well the first hydration shell geometries,\cite{Baer2010} so conclusions regarding the specific interactions and hydrate geometries between the \suldiox~and \wat~cannot be made here. It is notable that the same reorientation does not occur as strongly in the saturated system. The presence of the adsorbed \suldiox~layer apparently decreases the reorienting behavior likely because of the disrupting effect the higher \suldiox~concentration has on the water interactions at the surface. %Below this 5\angs~region under the surface, the \suldiox~resumes an isotropic orientation as it becomes fully solvated by bulk waters.
